Exploring the evolution of stellar rotation using Galactic kinematics

### 2020-04-17 paper/ms63.tex
adding jackie as coathor
commas in the abstract

\begin{abstract}
The rotational evolution of cool dwarfs is poorly constrained after $\sim$1-2
Gyr due to a lack of precise ages and rotation periods for old main-sequence
stars.
\racomment{
In this work, we use velocity dispersion as an age proxy, to reveal the
temperature-dependent rotational evolution of low-mass \kepler\ dwarfs and
demonstrate that kinematic ages could be a useful tool for calibrating
gyrochronology in the future.
}
% In this work, we use the velocity dispersions of groups of low-mass \kepler\
% dwarfs as an age proxy, to reveal the temperature-dependent rotational
% evolution of these stars, and demonstrate that kinematic ages could be a
% useful tool for calibrating gyrochronology in future.
We find that a linear gyrochronology model, calibrated to fit the
period--\teff\ relationship of the Praesepe cluster, does not apply to stars
older than around 1 Gyr.
Although late-K dwarfs spin more slowly than early-K dwarfs when they are
    young, at old ages we find that late-K dwarfs rotate at the {\it same
    rate} or faster than early-K dwarfs of the same age.
This result agrees qualitatively with semi-empirical models that vary the rate
of surface-to-core angular momentum transport as a function of time and mass.
It also aligns with recent observations of stars in the NGC 6811 cluster,
which indicate that the surface rotation rates of K dwarfs go through an epoch
of inhibited evolution.
We find that the oldest \kepler\ stars with measured rotation periods are
late-K and early-M dwarfs, indicating that these stars maintain spotted
surfaces and stay magnetically active longer than more massive stars.
Finally, based on their kinematics, we confirm that many rapidly rotating
GKM dwarfs are likely to be synchronized binaries.
\end{abstract}

>>>>

\author{Jacqueline K. Faherty}
\affiliation{Department of Astrophysics, American Museum of Natural History,
200 Central Park West, Manhattan, NY, USA}

\begin{abstract}
The rotational evolution of cool dwarfs is poorly constrained after $\sim$1-2
Gyr due to a lack of precise ages and rotation periods for old main-sequence
stars.
\racomment{
In this work we use velocity dispersion as an age proxy to reveal the
temperature-dependent rotational evolution of low-mass \kepler\ dwarfs, and
demonstrate that kinematic ages could be a useful tool for calibrating
gyrochronology in the future.
}
% In this work, we use the velocity dispersions of groups of low-mass \kepler\
% dwarfs as an age proxy, to reveal the temperature-dependent rotational
% evolution of these stars, and demonstrate that kinematic ages could be a
% useful tool for calibrating gyrochronology in future.
We find that a linear gyrochronology model, calibrated to fit the
period--\teff\ relationship of the Praesepe cluster, does not apply to stars
older than around 1 Gyr.
Although late-K dwarfs spin more slowly than early-K dwarfs when they are
    young, at old ages we find that late-K dwarfs rotate at the {\it same
    rate} or faster than early-K dwarfs of the same age.
This result agrees qualitatively with semi-empirical models that vary the rate
of surface-to-core angular momentum transport as a function of time and mass.
It also aligns with recent observations of stars in the NGC 6811 cluster,
which indicate that the surface rotation rates of K dwarfs go through an epoch
of inhibited evolution.
We find that the oldest \kepler\ stars with measured rotation periods are
late-K and early-M dwarfs, indicating that these stars maintain spotted
surfaces and stay magnetically active longer than more massive stars.
Finally, based on their kinematics, we confirm that many rapidly rotating
GKM dwarfs are likely to be synchronized binaries.
\end{abstract}
